---
title: "Group project cluster analysis"
author: "Daphney Loka"
date: "2026-04-11"
output: html_document
---

# Read datasets from csv files 

```{r}

# load datasets from csv files into R
if (!"readxl" %in% installed.packages()) install.packages("readxl")
library(readxl)

# import data
dta.pda <- read_excel("Loyalty Program Data_clean.xlsx", sheet = "clean1")

```

# Select variables & standardize

```{r}
# list the names of the variables in the dataset
names(dta.pda)

head(dta.pda)
```

```{r}
# choose only the BASE variables for the cluster analyses.

# standardize the data using function scale().
# it is important to standardize the data because the scale 
# of different variables may vary a lot. The variables with 
# larger scales will have more impact in the distance measure
dta.pda.clust <- scale(dta.pda[,2:9])

```


# Calculate distance matrix

```{r}
# calculate distance matrix using the standardized data
# the result is an upper triangular matrix with 9045 items (135*134/2)
dist.pda <- dist(dta.pda.clust, method = "euclidean") 

# show a small part of the matrix (distances among the first 6 cases)
as.matrix(dist.pda)[1:6,1:6]
## person 1 with person 2 -6
## person 2 with person 1, 3-6
```


# Figure out N of clusters

```{r tidy=TRUE}
# load the packages needed 
if (!"NbClust" %in% installed.packages()) install.packages("NbClust")
library(NbClust)
```

Software suggestion of N of clusters

```{r tidy=FALSE, fig.width=7, fig.height=5}

# run different indices, each proposes its optimal N of clusters
res.nbclust <- NbClust(
  data = dta.pda.clust, min.nc = 2, max.nc = 10, 
  method = "complete", index ="all"
)

```

```{r}
# summarize voting of best n of clusters
best_nc <- as.data.frame(t(res.nbclust$Best.nc), stringsAsFactors = TRUE)
best_nc$Number_clusters <- as.factor(best_nc$Number_clusters)
	
# drop invalid indices where Value_Index is NA, or N_Cluster is 0
best_nc_valid <- best_nc[!is.na(best_nc$Value_Index) & best_nc$Number_clusters != 0, ]
best_nc_valid$Number_clusters <- droplevels(best_nc_valid$Number_clusters)
best_nc_sum <- summary(best_nc_valid$Number_clusters)

# show histogram
df <- data.frame(N_Cluster = names(best_nc_sum), freq = best_nc_sum, stringsAsFactors = TRUE)
barplot(height = df$freq,
        names.arg = df$N_Cluster,
        xlab = "Number of clusters k",
        ylab = "Frequency among all indices",
        main = paste0("Optimal number of clusters - k = ", names(which.max(best_nc_sum))),
        col = "steelblue",
        border = "steelblue")

```

<br>

# 2 Groups PAM

Because of the weakness of K-means approach, we use a robust version of K-means based on medoids, Partitioning Around Medoids (PAM). The command is "pam()" instead of "kmeans()".

```{r tidy=TRUE}
# load the packages needed
if (!"cluster" %in% installed.packages()) install.packages("cluster")
library(cluster)
```

```{r tidy=TRUE}
# Conduct PAM cluster analysis
fgp.G2 <- pam(dist.pda, 2)

# Results: cluster membership
fgp.G2$clustering
```

```{r eval=FALSE}
# EXPORT results to csv file
write.csv(fgp.G2$clustering, "FGP_G2.csv")
```

<br>

# 7 Groups PAM

```{r tidy=TRUE}
# Conduct PAM cluster analysis
FGP.G7 <- pam(dist.pda, 7)

# Results: cluster membership
FGP.G7$clustering
```

```{r eval=FALSE}
# Export results to csv file
write.csv(FGP.G7$clustering, "FGP_G7.csv")
```

<br><br>

############ 
Clean 2 with purchase count for each merchant + SalesAmt instead of PointReward


```{r}
# import data
dta.fgp <- read_excel("Loyalty Program Data_clean.xlsx", sheet = "clean2")
```


# Select variables & standardize

```{r}
# list the names of the variables in the dataset
names(dta.fgp)

head(dta.fgp)
```

```{r}
# choose only the BASE variables for the cluster analyses.

# standardize the data using function scale().
# it is important to standardize the data because the scale 
# of different variables may vary a lot. The variables with 
# larger scales will have more impact in the distance measure
dta.fgp.clust <- scale(dta.fgp[,2:12])

```


# Calculate distance matrix

```{r}
# calculate distance matrix using the standardized data
# the result is an upper triangular matrix with 9045 items (135*134/2)
dist.fgp <- dist(dta.fgp.clust, method = "euclidean") 

# show a small part of the matrix (distances among the first 6 cases)
as.matrix(dist.fgp)[1:6,1:6]
## person 1 with person 2 -6
## person 2 with person 1, 3-6
```


# Figure out N of clusters

```{r tidy=TRUE}
# load the packages needed 
if (!"NbClust" %in% installed.packages()) install.packages("NbClust")
library(NbClust)
```

Software suggestion of N of clusters

```{r tidy=FALSE, fig.width=7, fig.height=5}

# run different indices, each proposes its optimal N of clusters
res.nbclust2 <- NbClust(
  data = dta.fgp.clust, min.nc = 2, max.nc = 10, 
  method = "complete", index ="all"
)

```

```{r}
# summarize voting of best n of clusters
best_nc2 <- as.data.frame(t(res.nbclust2$Best.nc), stringsAsFactors = TRUE)
best_nc2$Number_clusters <- as.factor(best_nc2$Number_clusters)
	
# drop invalid indices where Value_Index is NA, or N_Cluster is 0
best_nc_valid2 <- best_nc2[!is.na(best_nc2$Value_Index) & best_nc2$Number_clusters != 0, ]
best_nc_valid2$Number_clusters <- droplevels(best_nc_valid2$Number_clusters)
best_nc_sum2 <- summary(best_nc_valid2$Number_clusters)

# show histogram
df2 <- data.frame(N_Cluster2 = names(best_nc_sum2), freq2 = best_nc_sum2, stringsAsFactors = TRUE)
barplot(height = df2$freq,
        names.arg = df2$N_Cluster2,
        xlab = "Number of clusters k",
        ylab = "Frequency among all indices",
        main = paste0("Optimal number of clusters - k = ", names(which.max(best_nc_sum2))),
        col = "steelblue",
        border = "steelblue")

```

<br>

# 2 Groups PAM

Because of the weakness of K-means approach, we use a robust version of K-means based on medoids, Partitioning Around Medoids (PAM). The command is "pam()" instead of "kmeans()".

```{r tidy=TRUE}
# load the packages needed
if (!"cluster" %in% installed.packages()) install.packages("cluster")
library(cluster)
```

```{r tidy=TRUE}
# Conduct PAM cluster analysis
fgp.G2.1 <- pam(dist.fgp, 2)

# Results: cluster membership
fgp.G2.1$clustering
```

```{r eval=FALSE}
# EXPORT results to csv file
write.csv(fgp.G2.1$clustering, "FGP_G2_1.csv")
```

<br>


############ 
Clean 3 with everything


```{r}
# import data
dta.fgp1 <- read_excel("Loyalty Program Data_clean.xlsx", sheet = "clean3")
```


# Select variables & standardize

```{r}
# list the names of the variables in the dataset
names(dta.fgp1)

head(dta.fgp1)
```

```{r}
# choose only the BASE variables for the cluster analyses.

# standardize the data using function scale().
# it is important to standardize the data because the scale 
# of different variables may vary a lot. The variables with 
# larger scales will have more impact in the distance measure
dta.fgp.clust1 <- scale(dta.fgp1[,2:15])

```


# Calculate distance matrix

```{r}
# calculate distance matrix using the standardized data
# the result is an upper triangular matrix with 9045 items (135*134/2)
dist.fgp1 <- dist(dta.fgp.clust1, method = "euclidean") 

# show a small part of the matrix (distances among the first 6 cases)
as.matrix(dist.fgp1)[1:6,1:6]
## person 1 with person 2 -6
## person 2 with person 1, 3-6
```


# Figure out N of clusters

```{r tidy=TRUE}
# load the packages needed 
if (!"NbClust" %in% installed.packages()) install.packages("NbClust")
library(NbClust)
```

Software suggestion of N of clusters

```{r tidy=FALSE, fig.width=7, fig.height=5}

# run different indices, each proposes its optimal N of clusters
res.nbclust3 <- NbClust(
  data = dta.fgp.clust1, min.nc = 2, max.nc = 10, 
  method = "complete", index ="all"
)

```

```{r}
# summarize voting of best n of clusters
best_nc3 <- as.data.frame(t(res.nbclust3$Best.nc), stringsAsFactors = TRUE)
best_nc3$Number_clusters <- as.factor(best_nc3$Number_clusters)
	
# drop invalid indices where Value_Index is NA, or N_Cluster is 0
best_nc_valid3 <- best_nc3[!is.na(best_nc3$Value_Index) & best_nc3$Number_clusters != 0, ]
best_nc_valid3$Number_clusters <- droplevels(best_nc_valid3$Number_clusters)
best_nc_sum3 <- summary(best_nc_valid3$Number_clusters)

# show histogram
df3 <- data.frame(N_Cluster3 = names(best_nc_sum3), freq3 = best_nc_sum3, stringsAsFactors = TRUE)
barplot(height = df3$freq,
        names.arg = df3$N_Cluster3,
        xlab = "Number of clusters k",
        ylab = "Frequency among all indices",
        main = paste0("Optimal number of clusters - k = ", names(which.max(best_nc_sum3))),
        col = "steelblue",
        border = "steelblue")

```

<br>

# 3 Groups PAM

Because of the weakness of K-means approach, we use a robust version of K-means based on medoids, Partitioning Around Medoids (PAM). The command is "pam()" instead of "kmeans()".

```{r tidy=TRUE}
# load the packages needed
if (!"cluster" %in% installed.packages()) install.packages("cluster")
library(cluster)
```

```{r tidy=TRUE}
# Conduct PAM cluster analysis
fgp.G3 <- pam(dist.fgp1, 3)

# Results: cluster membership
fgp.G3$clustering
```

```{r eval=FALSE}
# EXPORT results to csv file
write.csv(fgp.G3$clustering, "FGP_G3.csv")
```

<br>



############ 
Clean 4 with everything (and edited age)


```{r}
# import data
dta.fgp1 <- read_excel("Loyalty Program Data_clean.xlsx", sheet = "clean4")
```


# Select variables & standardize

```{r}
# list the names of the variables in the dataset
names(dta.fgp1)

head(dta.fgp1)
```

```{r}
# choose only the BASE variables for the cluster analyses.

# standardize the data using function scale().
# it is important to standardize the data because the scale 
# of different variables may vary a lot. The variables with 
# larger scales will have more impact in the distance measure
dta.fgp.clust1 <- scale(dta.fgp1[,2:15])

```


# Calculate distance matrix

```{r}
# calculate distance matrix using the standardized data
# the result is an upper triangular matrix with 9045 items (135*134/2)
dist.fgp1 <- dist(dta.fgp.clust1, method = "euclidean") 

# show a small part of the matrix (distances among the first 6 cases)
as.matrix(dist.fgp1)[1:6,1:6]
## person 1 with person 2 -6
## person 2 with person 1, 3-6
```


# Figure out N of clusters

```{r tidy=TRUE}
# load the packages needed 
if (!"NbClust" %in% installed.packages()) install.packages("NbClust")
library(NbClust)
```

Software suggestion of N of clusters

```{r tidy=FALSE, fig.width=7, fig.height=5}

# run different indices, each proposes its optimal N of clusters
res.nbclust3 <- NbClust(
  data = dta.fgp.clust1, min.nc = 2, max.nc = 10, 
  method = "complete", index ="all"
)

```

```{r}
# summarize voting of best n of clusters
best_nc3 <- as.data.frame(t(res.nbclust3$Best.nc), stringsAsFactors = TRUE)
best_nc3$Number_clusters <- as.factor(best_nc3$Number_clusters)
	
# drop invalid indices where Value_Index is NA, or N_Cluster is 0
best_nc_valid3 <- best_nc3[!is.na(best_nc3$Value_Index) & best_nc3$Number_clusters != 0, ]
best_nc_valid3$Number_clusters <- droplevels(best_nc_valid3$Number_clusters)
best_nc_sum3 <- summary(best_nc_valid3$Number_clusters)

# show histogram
df3 <- data.frame(N_Cluster3 = names(best_nc_sum3), freq3 = best_nc_sum3, stringsAsFactors = TRUE)
barplot(height = df3$freq,
        names.arg = df3$N_Cluster3,
        xlab = "Number of clusters k",
        ylab = "Frequency among all indices",
        main = paste0("Optimal number of clusters - k = ", names(which.max(best_nc_sum3))),
        col = "steelblue",
        border = "steelblue")

```

<br>

# 3 Groups PAM

Because of the weakness of K-means approach, we use a robust version of K-means based on medoids, Partitioning Around Medoids (PAM). The command is "pam()" instead of "kmeans()".

```{r tidy=TRUE}
# load the packages needed
if (!"cluster" %in% installed.packages()) install.packages("cluster")
library(cluster)
```

```{r tidy=TRUE}
# Conduct PAM cluster analysis
fgp.G3 <- pam(dist.fgp1, 3)

# Results: cluster membership
fgp.G3$clustering
```

```{r eval=FALSE}
# EXPORT results to csv file
write.csv(fgp.G3$clustering, "FGP_G3_1.csv")
```

<br>


